### 2016-12-01 chapter1.tex
Incorperated superdarnmap figure
Small corrections made throughout to references

%Figures
%- Map of radars (Northern and southern hemisphere)
%- Example convection map
%- Example ISR profile
%- Example MSISE profile?
%- Example of IGRF?
%- How does radar work diagram
%- Pulse sequence diagram

@@

%BENIFITS OF MULTIPULSE: Ponomarenko and Waters, 2006
Instead of waiting for each individual signal to be recieved before transmitting the next, ionospheric radars typically employ a multipulse mode. In a multipulse mode, the radar transmits a series of pules with different time intervals or lags between them.  This can introduce complications when backscatter from different pulses at different ranges is received by the radar at the same time, referred to as cross-range interference, but these effects can be mitigated using correlation techniques.  The smallest lag between two pulses is known as the multi-pulse increment, \(\tau\).  All other lags are integer multiples of \(\tau\), which allows the calculation of the corresponding lags of the complex autocorrelation function (ACF).  The ACF is used to find the spectral characteristics of the backscatter, from which the LoS velocity, spectral width, and backscatter power can be obtained.  Overall, multi-pulse techniques are generally considered far more appropriate for ionospheric studies than single pulse \citep{Farley1972,Greenwald1983,Greenwald1985,Barthes1998,Ponomarenko2006}.

@@

%- How to measure power, spectral width, LoS velocity
%	- What are these things?
%- Hanuise, 1993b
%- Baker, 1995

@@

\begin{figure}
	\includegraphics[width=\textwidth]{superdarnmap.pdf}
	\caption{FoVs of all operational SuperDARN radars in both the northern (left) and southern (right) hemispheres.  Polar latitude radars are shown with the red outlines.  The radars at Rankin Inlet (northern hemisphere) and McMurdo Station (southern hemisphere) are shaded in red because these two instruments are particularly important in the following studies.  Lines of constant magnetic latitude are shown in yellow at \(\Lambda = 40\deg, 60\deg, 80\deg\) in the northern hemisphere and at \(\Lambda = -40\deg,-60\deg,-80\deg\) in the southern hemisphere.}
	\label{fig:superdarnmap}
\end{figure}

@@

One of the original purposes of SuperDARN was to produce maps of the plasma convection patterns in the polar caps using 2D velocity vectors.  Originally this was accomplished by considering two radars with overlapping FoVs.  If both recieved backscatter from the same scattering volume, two LoS velocities could be found for that scattering volume, which could be combined to find a 2D velocity vector \citep{Ruohoniemi1989}.  The modern method for creating convection maps involves taking all data recorded by all radars in a particular hemisphere and finding the best fit to a 2D ionospheric electrostatic potential using Legendre functions \citep{Ruohonomiemi1998}.  For convection maps, only data from the F region is considered.  If there is not enough velocity data to confine the fit sufficiently, the measured data points are supplimented with a statistical model based on solar wind IMF conditions \textbf{CITE MCWILLIAMS???}.

@@

Analysis of the returned power spectrum reveals the electron density, electron and ion temperatures, and LoS ion velocity \citep{Evans1969,Rishbeth1985,Nicolls2007}.  Using a model of chemical composition at different altitudes, the average ion mass can also be found.  In addition, 3D convection velocities and electric field vectors can be derived from LoS ion velocity measurements in some of the more modern ISRs by assuming both \(\vec{E}\) and \(\vec{V}_E\) are constant across the radar's FoV \citep{Heinselman2008}.  Each instrument must be calibrated with known electron density measurements to account for noise and system constants \citep{Nicolls2007}.  This is done using the plasma frequency measured from either the plasma line of the ISR spectra during summer daytime period or ionosonde measurements for all other times \citep{Bachivan2010,Themens2014}.

@@

The IRI model can produce outputs of electron density, temperatures of electrons, ions, and neutrals, ion composition (O+, H+, He+, O2+, NO+, N+), and total electron content (TEC).  The model requires the input of a location (latitude, longitude, and altitude) and time (year, date, and time).  Additionally, the model gives the option to choose between two F peak models and three bottomside thickness models.  There are a variety of other optional input parameters such as sunspot number, ionospheric index, and F10.7 radio flux which can help make the output parameters more accurate for a particular situation.  Internally, the model has the additional capability of producing profiles of any of the output parameters in either space or time.  This is particularly significant for height profiles, and allows the peak heights and densities of the E and F region to be calculated.

Although the IRI model is very useful for providing background plasma density conditions where no instruments are available to make real measurements, it is important to remember that it is a model and therefor only represents the average conditions.  Although it is fairly accurate at mid-latitudes \citep{Coisson2006,Bilitza2012}, it does not necessarily represent variations within the polar cap well \citep{Themens2014,Makarevich2015}.  In particular, the height of the F-region peak tends to be underestimated in daytime and the bottomside thickness does not show realistic seasonal and diurnal variation \citep{Themens2014}.  Despite this, the IRI model is still a very important tool for establishing a density profile at any time in any location in the ionosphere.  And example of an altitude profile of plasma density provided by the IRI model is shown in Figure \ref{fig:densprofile}.  The displayed profiles shows how the electron density changes with altitude at local noon (day) and midnight (night) at \(75\deg\) N, \(0\deg\) E, geographic.

@@

\section{Theory and Observations}
Much work has been done over the past several decades to try to understand the complicated picture of plasma structuring in the polar cap, both from theoretical analysis of plasma dispersion relations, simulations of plasma dynamics, and observations of plasma structuring using a variety of instruments.

@@

There have been at least 6 mechanisms proposed for how polar patches form on the dayside.
\begin{enumerate}
	\item Discrete changes in solar wind parameters, usch as IMF, density, speed, and pressure \citep{Sojka1994}
	\item Average flow patterns varying in time \citep{Anderson1988}
	\item Transient magnetopause reconnection \citep{Lockwook1992b}
	\item Plasma production by cusp particle precipitation \citep{Rodger1994,Millward1999}
	\item Plasma flow jet channels that cut continuous streaches of plasma into segments 	\citep{Valladares1998}
	\item Alfven wave coupling \citep{Prikryl1999}
\end{enumerate}
Although particle precipitation has been shown to create enhanced plasma density in the cusp region \citep{Roger1994}, it is unreasonable for this mechanism to account for the largest density enhancements observed in polar patches.  These high densities must originate from the resivor of solar illuminated plasma on the dayside.  Additionally, a mechanism is still required to separate patches from the cusp region.  High speed flow channels can "cut" segments off of a tongue of high-density plasma that has been pulled into the polar cap by convection \citep{Valladares1994,Valladares1998}.  Models have also shown changing large scale convection patterns quickly in time does produce density structures similar to patches \citep{Anderson1988}.  However, the mechanism that seems to be dominant for the majority of polar patches observed is transient magnetic reconnection \citep{Carlson2012}.

@@

@@

>>>>

@@

Instead of waiting for each individual signal to be recieved before transmitting the next, ionospheric radars typically employ a multipulse mode. In a multipulse mode, the radar transmits a series of pules with different time intervals or lags between them.  This can introduce complications when backscatter from different pulses at different ranges is received by the radar at the same time, referred to as cross-range interference, but these effects can be mitigated using correlation techniques.  The smallest lag between two pulses is known as the multi-pulse increment, \(\tau\).  All other lags are integer multiples of \(\tau\), which allows the calculation of the corresponding lags of the complex autocorrelation function (ACF).  The ACF is used to find the spectral characteristics of the backscatter, from which the LoS velocity, spectral width, and backscatter power can be obtained.  Overall, multi-pulse techniques are generally considered far more appropriate for ionospheric studies than single pulse \citep{Farley1972,Greenwald1983,Greenwald1985,Barthes1998,Ponomarenko2006}.

@@

@@

\begin{figure}
	\includegraphics[width=\textwidth,angle=180]{superdarnmap.pdf}
	\caption{FoVs of all operational SuperDARN radars in both the northern (left) and southern (right) hemispheres.  Polar latitude radars are shown with the red outlines.  The radars at Rankin Inlet (northern hemisphere) and McMurdo Station (southern hemisphere) are shaded in red because these two instruments are particularly important in the following studies.  Lines of constant magnetic latitude are shown in yellow at \(\Lambda = 40\deg, 60\deg, 80\deg\) in the northern hemisphere and at \(\Lambda = -40\deg,-60\deg,-80\deg\) in the southern hemisphere.}
	\label{fig:superdarnmap}
\end{figure}

@@

One of the original purposes of SuperDARN was to produce maps of the plasma convection patterns in the polar caps using 2D velocity vectors.  Originally this was accomplished by considering two radars with overlapping FoVs.  If both recieved backscatter from the same scattering volume, two LoS velocities could be found for that scattering volume, which could be combined to find a 2D velocity vector \citep{Ruohoniemi1989}.  The modern method for creating convection maps involves taking all data recorded by all radars in a particular hemisphere and finding the best fit to a 2D ionospheric electrostatic potential using Legendre functions \textbf{CITE???}.  For convection maps, only data from the F region is considered.  If there is not enough velocity data to confine the fit sufficiently, the measured data points are supplimented with a statistical model based on solar wind IMF conditions \textbf{CITE MCWILLIAMS???}.

@@

Analysis of the returned power spectrum reveals the electron density, electron and ion temperatures, and LoS ion velocity \citep{Evans1969,Rishbeth1985,Nicolls2007a}.  Using a model of chemical composition at different altitudes, the average ion mass can also be found.  In addition, 3D convection velocities and electric field vectors can be derived from LoS ion velocity measurements in some of the more modern ISRs by assuming both \(\vec{E}\) and \(\vec{V}_E\) are constant across the radar's FoV \citep{Heinselman2008}.  Each instrument must be calibrated with known electron density measurements to account for noise and system constants \citep{Nicolls2007a}.  This is done using the plasma frequency measured from either the plasma line of the ISR spectra during summer daytime period or ionosonde measurements for all other times \citep{Bahcivan2010,Themens2014}.

@@

The IRI model can produce outputs of electron density, temperatures of electrons, ions, and neutrals, ion composition (O+, H+, He+, O2+, NO+, N+), and total electron content (TEC).  The model requires the input of a location (latitude, longitude, and altitude) and time (year, date, and time).  Additionally, the model gives the option to choose between two F peak models and three bottomside thickness models.  There are a variety of other optional input parameters such as sunspot number, ionospheric index, and F10.7 radio flux which can help make the output parameters more accurate for a particular situation.  Internally, the model has the additional capability of producing profiles of any of the output parameters in either space or time.  This is particularly significant for height profiles, and allows the peak heights and densities o
f the E and F region to be calculated.

Although the IRI model is very useful for providing background plasma density conditions where no instruments are available to make real measurements, it is important to remember that it is a model and therefor only represents the average conditions.  Although it is fairly accurate at mid-latitudes \citep{Coisson2006,Bilitza2012}, it does not necessarily represent variations within the polar cap well \citep{Themens2014,Makarevich2015}.  In particular, the height of the F-region peak tends to be underestimated in daytime and the bottomside thickness does not show realistic seasonal and diurnal variation \citep{Themens2014}.  Despite this, the IRI model is still a very important tool for establishing a density profile at any time in any location in the ionosphere.  And example of an altitude profile of plasma density provided by the IRI model is shown in Figure \ref{fig:densprofile}.  The displayed profiles shows how the electron density changes with altitude at local noon (day) and midnight (night) on January 1, 2007 at \(75\deg\) N, \(0\deg\) E, geographic.

@@

\section{Theory and Observations}
Much work has been done over the past several decades to try to understand the complicated picture of plasma structuring in the polar cap, including theoretical analysis of plasma dispersion relations, simulations of plasma dynamics, and observations of plasma structuring using a variety of instruments.

@@

There have been at least 6 mechanisms proposed for how polar patches form on the dayside.
\begin{enumerate}
	\item Discrete changes in solar wind parameters, usch as IMF, density, speed, and pressure \citep{Sojka1994}
	\item Average flow patterns varying in time \citep{Anderson1988}
	\item Transient magnetopause reconnection \citep{Lockwook1992b}
	\item Plasma production by cusp particle precipitation \citep{Rodger1994,Millward1999}
	\item Plasma flow jet channels that cut continuous streaches of plasma into segments 	\citep{Valladares1998}
	\item Alfven wave coupling \citep{Prikryl1999}
\end{enumerate}
Although particle precipitation has been shown to create enhanced plasma density in the cusp region \citep{Roger1994}, it is unreasonable for this mechanism to account for the largest density enhancements observed in polar patches.  These high densities must originate from the resivor of solar illuminated plasma on the dayside.  Additionally, a mechanism is still required to separate patches from the cusp region.  High speed flow channels can ``cut'' segments off of a tongue of high-density plasma that has been pulled into the polar cap by convection \citep{Valladares1994,Valladares1998}.  Models have also shown changing large scale convection patterns quickly in time does produce density structures similar to patches \citep{Anderson1988}.  However, the mechanism that seems to be dominant for the majority of polar patches observed is transient magnetic reconnection \citep{Carlson2012}.

@@

\begin{figure}
	\includegraphics[width=\textwidth]{magnetosphere.pdf}
	\caption{Diagram of the interactions between the IMF, magnetosphere, and polar cap ionosphere.  Closed field lines in the Earth's magnetosphere are shown in grey while open field lines are light blue.  The polar cap open-closed boundary is in green.  Magnetic field lines that are part of the IMF are shown in dark blue.  Magnetic reconnection occurs along the X-line (red) on the dayside of the magnetopause (dashed line).  The X-line can be mapped back along magnetic field lines (orange) to the ionosphere, where their projection is the merging gap (pink dotted line).  Durring reconnection events, the open-closed boundary expands equatorwards as more open field lines are formed and there is plasma flow across the merging gap.}
	\label{fig:magnetosphere}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=\textwidth]{patch_formation.pdf}
	\caption{Diagram of the polar patch formation through transient magnetopause reconnection.  The open-closed boundary is the green line.  Black contours represent plasma flow.  Highly ionized (dense) daytime plasma is shown by the light orange regions.  The burst of reconnection is occurring in Figures \ref{fig:patch_formation}a--\ref{fig:patch_formation}c but stops in Figures \ref{fig:patch_formation}d--\ref{fig:patch_formation}f.}
	\label{fig:patch_formation}
\end{figure}
